## Analyisis: PCA for monthly changes in implied volatilties

```{r implied_volatilities_distribution, include = FALSE, eval = FALSE}
swaption_volatilities_long <- swaption_volatilities |>
  pivot_longer(
    cols = -Timestamp,
    names_to = "name",
    values_to = "value"
  )

# calculates plug-in bandwidth of Silverman (1986)
# data: vector of data
# returns bandwidth
bandwidth_silverman <- function(data) {
  return(0.9 * min(sd(data), IQR(data) / 1.34) * length(data)^(-1 / 5))
}

for (colname in colnames(swaption_volatilities)[-1]) {
  data <- swaption_volatilities_long |>
    filter(name %in% colname)
  data_vector <- as.numeric(data$value)
  kde <- density(data_vector, bw = bandwidth_silverman(data_vector), from = min(data_vector), to = max(data_vector))
  # data frame with visualization data
  vizdata <- tibble(
    x = kde$x,
    y = kde$y
  )

  plot <- ggplot(data = data, aes(x = value)) +
    geom_histogram() +
    # geom_histogram(data = data |> filter(Timestamp >= ymd("2020-01-01")), fill = "orange", alpha = 0.2) +
    geom_histogram(data = data |> filter(Timestamp < ymd("2017-01-01")), fill = "green", alpha = 0.2) +
    geom_line(
      data = vizdata,
      mapping = aes(x = x, y = y),
      linewidth = 0.8, color = "blue"
    ) +
    labs(
      title = "Histogram of daily swaption implied volatilities",
      subtitle = paste0("Swaption ", colname, " from 2/7/2013 to 7/15/2024"),
      x = colname,
      y = "density"
    )
  print(plot)
}
```

**PCA and annualization for weekly changes in implied volatilities (start date: 2013-02-01)**
```{r}
swaption_volatilities_w_rel %>%
  pivot_longer(
    cols = -Timestamp,
    names_to = "name",
    values_to = "value"
  ) %>%
  group_by(name) %>%
  summarize(
    min = min(value, na.rm = TRUE),
    quantile1 = quantile(value, 0.01, na.rm = TRUE),
    quantile25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    mean = mean(value, na.rm = TRUE),
    quantile75 = quantile(value, 0.75, na.rm = TRUE),
    quantile99 = quantile(value, 0.99, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pander()
pander(swaption_volatilities_d.summary)

swaption_volatilities_w_rel <- swaption_volatilities_by_date_and_interval(swaption_volatilities_d, ymd("2013-02-01"), "week")

pca <- prcomp(swaption_volatilities_w_rel[, -1], center = TRUE, scale = FALSE)

pca$rotation[, "PC2"] <- -pca$rotation[, "PC2"]

fviz_eig(pca)
fviz_pca_ind(pca, label = "none")

get_eigenvalue(pca) |> pander()

pca_annualized <- principal_component_viz(pca, swaption_volatilities_w_rel, 3, with_plots = TRUE, is_weekly = TRUE)

###############################################################################
# Transform first three 99.5% quantile principal component scores back to
# original coordinates
###############################################################################
pca_swaption_volatilities_copula <- t(t(pca_annualized$joint_quantile_copula$joint_quantile %*% t(pca$rotation[, 1:3])) + pca$center) |> as_tibble()
plotly_3d(
  pca_swaption_volatilities_copula |> as.numeric(),
  "99.5% quantile estimates using PCA and annualization [Copula]"
)

pca_swaption_volatilities_ar1 <- t(t(pca_annualized$joint_quantile_ar$joint_quantile %*% t(pca$rotation[, 1:3])) + pca$center) |> as_tibble()
plotly_3d(
  pca_swaption_volatilities_ar1 |> as.numeric(),
  "99.5% quantile estimates using PCA and annualization [AR(1)]"
)

vizdata <- matrix(c(pca_swaption_volatilities_copula |> as.numeric(), pca_swaption_volatilities_ar1 |> as.numeric()),
  ncol = 2, byrow = FALSE
)
colnames(vizdata) <- c("Copula annualization", "AR(1) annualization")
plotly_3d(vizdata, "99.5% quantile estimates using PCA and annualization")
```
```{r}
plot <- ggplot(tibble(x = pobs(pca$x[1:597, "PC1"]), y = pobs(pca$x[2:598, "PC1"])), aes(x, y)) +
  geom_point() +
  labs(
    title = "Observations of lag_0 and lag_1 scores",
    subtitle = "598 weekly first principal component scores",
    x = "marginal uniform distribution 1",
    y = "marginal uniform distribution 2"
  )
print(ggMarginal(plot, type = "density"))
```


```{r}
plots_histogram <- list()
plots_projection <- list()
plots_acf <- list()
plots_pacf <- list()

for (order_number in 1:3) {
  order_text <- ""
  if (order_number == 1) {
    order_text <- "first"
  } else if (order_number == 2) {
    order_text <- "second"
  } else if (order_number == 3) {
    order_text <- "third"
  }

  df_pc_projection <- tibble(
    timestamp = swaption_volatilities_w_rel$Timestamp,
    projection = NA
  )
  for (i in 1:nrow(df_pc_projection)) {
    df_pc_projection$projection[i] <- sum(
      swaption_volatilities_w_rel[i, -1] * pca$rotation[, paste0("PC", order_number)]
    )
  }

  pc_scores_distr <- fitdistr(df_pc_projection$projection, "normal")$estimate
  print(pc_scores_distr)
  plot_qq_norm <- ggplot(df_pc_projection, aes(sample = projection)) +
    stat_qq(distribution = qnorm, dparams = c(pc_scores_distr["mean"], pc_scores_distr["sd"])) +
    stat_qq_line(distribution = qnorm, dparams = c(pc_scores_distr["mean"], pc_scores_distr["sd"])) +
    labs(
      title = "QQ-plot: Normal distribution and scores",
      subtitle = "Quantiles of observed scores vs fitted normal distribution",
      x = "theoretical quantile",
      y = "sample quantile"
    )

  pc_scores_distr <- fitdistr(df_pc_projection$projection, "t")$estimate
  print(pc_scores_distr)
  print(fitdistr(df_pc_projection$projection, "t")$loglik)

  pc_scores_distr_tibble <- tibble(
    x = seq(min(df_pc_projection$projection), max(df_pc_projection$projection), length.out = 500),
    # density of fitted t distribution (see formula for density of scaled t distribution)
    y = dt((x - pc_scores_distr["m"]) / pc_scores_distr["s"], df = pc_scores_distr["df"]) / pc_scores_distr["s"]
  )
  print(paste0("pcscoresdistrs", pc_scores_distr["s"]))
  plot_qq_t <- ggplot(df_pc_projection, aes(sample = (projection - pc_scores_distr["m"]) / pc_scores_distr["s"])) +
    stat_qq(distribution = qt * pc_scores_distr["s"] + pc_scores_distr["m"], dparams = pc_scores_distr["df"]) +
    stat_qq_line(distribution = qt, dparams = pc_scores_distr["df"]) +
    labs(
      title = "QQ-plot: t distribution and scores",
      subtitle = "Quantiles of observed scores vs fitted scaled t distribution",
      x = "theoretical quantile",
      y = "sample quantile"
    )
  print(plot_qq_norm | plot_qq_t)

  subtitle <- glue(
    '<span style="color:#FF0000">**Kernel density estimate**</span>, ',
    '<span style="color:#0000FF">**scaled <i>t</i> distribution** fitted using MLE</span>',
  )

  plot_histogram <- ggplot(data = df_pc_projection, aes(x = projection)) +
    geom_histogram(aes(y = ..density..)) +
    geom_density(colour = "red", linewidth = 0.8) +
    geom_line(data = pc_scores_distr_tibble, aes(x, y), color = "blue", linewidth = 0.8) +
    # annotate(geom = "text", x = -Inf, y = Inf, hjust=0, vjust=1,
    #          label = TeX(paste0("$\\hat{\\mu}$ = ", formatC(mean(df_pc_projection$projection), digits=4),
    #             ", $\\hat{\\sigma}^2$ = ", formatC(var(df_pc_projection$projection)|>sqrt(), digits=4),
    #             ", $\\hat{\\gamma}$ = ", formatC(skewness(df_pc_projection$projection), digits=4),
    #             ", $\\hat{\\kappa}$ = ", formatC(kurtosis(df_pc_projection$projection), digits=4)))) +
    labs(
      title = paste0("Histogram of ", order_text, " principal component scores"),
      subtitle = subtitle,
      x = "principal component score",
      y = "density"
    ) +
    theme(
      plot.subtitle = element_markdown(),
      legend.position = "none"
    )
  plots_histogram[[order_number]] <- plot_histogram

  plot_projection <- ggplot() +
    geom_line(data = df_pc_projection, mapping = aes(timestamp, projection)) +
    annotate(geom = "rect", xmin = ymd("2015-05-01"), xmax = ymd("2015-06-01"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2) +
    annotate(geom = "rect", xmin = ymd("2020-03-01"), xmax = ymd("2020-04-01"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2) +
    labs(
      title = paste0("Scores of ", order_text, " principal component"),
      x = "time",
      y = ""
    )
  plots_projection[[order_number]] <- plot_projection

  acf <- acf(df_pc_projection$projection, lag.max = 50, plot = FALSE)
  df_acf <- with(acf, data.frame(lag, acf))
  conf.level <- 0.95
  ciline <- qnorm((1 - conf.level) / 2) / sqrt(length(df_pc_projection$projection))
  plot_acf <- ggplot(data = df_acf, mapping = aes(x = lag, y = acf)) +
    geom_segment(mapping = aes(xend = lag, yend = 0), size = 1.5, colour = "gray30") +
    labs(
      title = paste0("Autocorrelation function of ", order_text, " principal component"),
      x = "lag (weeks)",
      y = ""
    )

  pacf <- pacf(df_pc_projection$projection, lag.max = 50, plot = FALSE)
  pacf <- pacf(df_pc_projection$projection, lag.max = 50, plot = FALSE)
  pdf_acf <- with(pacf, data.frame(lag, acf))

  plot_pacf <- ggplot(data = pdf_acf, mapping = aes(x = lag, y = acf)) +
    geom_hline(aes(yintercept = 0)) +
    geom_segment(mapping = aes(xend = lag, yend = 0), size = 1.5, colour = "gray30") +
    geom_hline(aes(yintercept = ciline), linetype = 2, color = "darkblue") +
    geom_hline(aes(yintercept = -ciline), linetype = 2, color = "darkblue") +
    labs(
      title = paste0("Sample partial autocorrelation function of ", order_text, " principal component"),
      x = "lag (weeks)",
      y = ""
    )
  plots_acf[[order_number]] <- plot_acf
  plots_pacf[[order_number]] <- plot_pacf
}

(plots_histogram[[1]] | plots_projection[[1]]) / (plots_histogram[[2]] | plots_projection[[2]]) / (plots_histogram[[3]] | plots_projection[[3]])

(plots_acf[[1]] | plots_pacf[[1]]) / (plots_acf[[2]] | plots_pacf[[2]]) / (plots_acf[[3]] | plots_pacf[[3]])
```

```{r}
swaption_volatilities_m_rel <- swaption_volatilities_by_date_and_interval(swaption_volatilities_d, ymd("2013-02-01"), "month")
pca <- prcomp(swaption_volatilities_m_rel[, -1], center = TRUE, scale = FALSE)

fviz_eig(pca)
fviz_pca_ind(pca, label = "none")

get_eigenvalue(pca) |> pander()

pca_annualized <- principal_component_viz(pca, swaption_volatilities_m_rel, 3, with_plots = TRUE, is_weekly = FALSE)

###############################################################################
# Transform first three 99.5% quantile principal component scores back to
# original coordinates
###############################################################################
pca_swaption_volatilities_copula <- t(t(pca_annualized$joint_quantile_copula$joint_quantile %*% t(pca$rotation[, 1:3])) + pca$center) |> as_tibble()
plotly_3d(
  pca_swaption_volatilities_copula |> as.numeric(),
  "99.5% quantile estimates using PCA and annualization [Copula]"
)

pca_swaption_volatilities_ar1 <- t(t(pca_annualized$joint_quantile_ar$joint_quantile %*% t(pca$rotation[, 1:3])) + pca$center) |> as_tibble()
plotly_3d(
  pca_swaption_volatilities_ar1 |> as.numeric(),
  "99.5% quantile estimates using PCA and annualization [AR(1)]"
)

vizdata <- matrix(c(pca_swaption_volatilities_copula |> as.numeric(), pca_swaption_volatilities_ar1 |> as.numeric()),
  ncol = 2, byrow = FALSE
)
colnames(vizdata) <- c("Copula annualization", "AR(1) annualization")
plotly_3d(vizdata, "99.5% quantile estimates using PCA and annualization")
```
The two surface plots above show the 99.5% quantile estimates using PCA and annualization for the annual change in implied volatilities. The surface plot obtained using aggregated AR(1) simulations is closely related to the one in *Observed annual change in implied volatilities*.

```{r}
row_index <- which(swaption_volatilities_w_rel$Timestamp == ymd("2015-05-05"))
pca_projection_example <- t(pca$x[row_index, 1:3] %*% t(pca$rotation[, 1:3])) + pca$center

surfaces <- matrix(c(c(swaption_volatilities_w_rel[row_index, -1]), c(pca_projection_example)), ncol = 2, byrow = FALSE)
colnames(surfaces) <- c("Market surface", "PCA surface")

plotly_3d(
  surfaces,
  "", "rel. change in implied volatility"
)
```

**Stability of annualization methods (start date: 2013-02-01)**
```{r, stability_annualization_methods}
annual_quantiles <- tibble()

# swaption_volatilities_m_rel = swaption_volatilities_by_date_and_interval(swaption_volatilities_d, ymd("2013-02-01"), 'month')
# pca = prcomp(swaption_volatilities_m_rel[,-1], center = TRUE, scale = FALSE)
# for (i in 1:20) {
#   quantiles = c()
#   for (principal_component in 1:3) {
#     ret_val = principal_component_viz(pca, swaption_volatilities_m_rel, principal_component, conf_level = 0.995, with_plots = FALSE)
#     quantiles = c(quantiles, ret_val$upper_quantiles)
#   }
#   annual_quantiles = rbind(annual_quantiles, quantiles)
# }
#
# colnames(annual_quantiles) = c('pc1_quantile_copula', 'pc1_quantile_ar', 'pc2_quantile_copula', 'pc2_quantile_ar', 'pc3_quantile_ar', 'pc3_quantile_copula')
# save(annual_quantiles, file = 'annual_quantiles_stability.RData')
load("annual_quantiles_stability.RData")

# histogram of quantiles of principal component scores with different start date
ggplot(data = gather(annual_quantiles), aes(x = value, y = factor(""))) +
  geom_violin() +
  facet_wrap(~ factor(key, c("pc1_quantile_copula", "pc2_quantile_copula", "pc3_quantile_copula", "pc1_quantile_ar", "pc2_quantile_ar", "pc3_quantile_ar")), scales = "free") +
  labs(
    title = paste0("99.5% quantiles of 20 annual principal component simulated scores"),
    subtitle = paste0("Groups by principal component and annualization approach"),
    x = "99.5% quantile",
    y = "density"
  )
```
20 repeated simulations of the 99.5% quantile estimators of annualized principal component scores confirm that the estimators are stable. However, the values of the quantile estimators differ for the two annualization methods.

**PCA and annualization  for monthly changes in implied volatilities (rolling start dates)**

```{r}
annual_quantiles <- tibble()

start_dates <- (swaption_volatilities_d |> arrange(Timestamp))$Timestamp[seq(1, 100, length.out = 6)]
for (i in 1:length(start_dates)) {
  swaption_volatilities_w_rel <- swaption_volatilities_by_date_and_interval(swaption_volatilities_d, start_dates[i], "week")
  pca <- prcomp(swaption_volatilities_w_rel[, -1], center = TRUE, scale = FALSE)
  pca$rotation[, "PC2"] <- -pca$rotation[, "PC2"]
  # for (principal_component in 1:3) {
  ret_val <- principal_component_viz(pca, swaption_volatilities_w_rel, 3, with_plots = FALSE, is_weekly = TRUE)
  quantiles <- c(ret_val$joint_quantile_copula$joint_quantile, ret_val$joint_quantile_ar$joint_quantile)
  # }
  annual_quantiles <- rbind(annual_quantiles, quantiles)
}

colnames(annual_quantiles) <- c("pc1_quantile_copula", "pc1_quantile_ar", "pc2_quantile_copula", "pc2_quantile_ar", "pc3_quantile_ar", "pc3_quantile_copula")
save(annual_quantiles, file = "annual_quantiles.RData")
load("annual_quantiles.RData")

# histogram of quantiles of principal component scores with different start date
ggplot(data = gather(annual_quantiles), aes(x = value, y = factor(""))) +
  geom_violin() +
  geom_boxplot(width = 0.4) +
  facet_wrap(~ factor(key, c("pc1_quantile_copula", "pc2_quantile_copula", "pc3_quantile_copula", "pc1_quantile_ar", "pc2_quantile_ar", "pc3_quantile_ar")), scales = "free") +
  labs(
    title = paste0("99.5% quantiles of scores for various start dates"),
    subtitle = paste0("Groups by principal component and annualization approach"),
    x = "99.5% quantile of principal component scores",
    y = ""
  )
```
The 99.5% quantile estimators vary slightly for reference dates +/- 30 days.

```{r, eval = FALSE, include = FALSE}
tmp <- swaption_volatilities_d |>
  filter(Timestamp == ymd("2023-12-29"))
plotly_3d(as.numeric(tmp[, -1]), "Swaption volatility surface on 2023-12-29")
```