---
title: "Swaptions: Dynamics of implied volatility surfaces"
author: "David Neuhäusler"
date: "February 2025"
site: bookdown::bookdown_site
documentclass: book
description: |
  R scripts for the master's thesis 'Dynamics of implied swaption volatility  surfaces and implications on life insurers under Solvency II' at Ulm University
link-citations: yes
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE
)
```

```{r load_packages, include=TRUE}
library(MASS)
library(tidyverse) # functions for data visualization
library(factoextra) # functions for multivariate data analyses
library(plotly) # plot library
library(lubridate) # date helper functions
library(patchwork) # simple layouts for ggplot2
library(ggpubr) # simple layouts for ggplot2
library(ggExtra) # extend ggplot2 toolset
library(ggpubr) # customize ggplot2 plots
library(ggtext) # simple Markdown and HTML rendering for ggplot2
library(glue) # R expression embedding for strings
library(formattable) # formatter functions
library(latex2exp) # use LaTeX in plots
library(tseries) # time series analysis and computational finance
library(pracma) # practical numerical math functions
library(zoo) # time series utils
library(pander) # output formatting of tibbles and data frames
library(timetk) # filter date and time expressions
library(dplyr) # data manipulation
library(ggcorrplot) # plot covariance matrices
library(copula) # copulas
library(quantreg) # quantile regression
library(Sim.DiffProc)
library(mice) # imputation of missing values
```

```{r configure_theme, include=FALSE}
theme_set(theme_minimal() %+replace%
  theme(
    # plot title
    plot.title = element_text(
      size = rel(1.3), face = "bold",
      margin = margin(0, 0, 5, 0), hjust = 0
    ),
    # plot panel
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    # axes
    axis.title = element_text(size = rel(0.85), face = "bold"),
    axis.text = element_text(size = rel(0.85), face = "bold"),
    axis.line = element_line(color = "black"),
    # legend
    legend.title = element_text(size = rel(0.85), face = "bold"),
    legend.text = element_text(size = rel(0.70), face = "bold"),
    legend.key = element_rect(fill = "transparent", colour = NA),
    legend.key.size = unit(1.5, "lines"),
    legend.background = element_rect(fill = "transparent", colour = NA),
    # plot facets
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text = element_text(
      size = rel(0.85), face = "bold", color = "white",
      margin = margin(5, 0, 5, 0)
    )
  ))
colors <- c(
  "#000000", "#555555", "#999999",
  "#0072B2", "#009E73", "#E69F00", "#D55E00",
  "#F0E442", "#CC79A7", "#56B4E9"
)
names(colors) <- c(
  "black", "gray", "lightgray", "blue", "green",
  "orange", "red", "yellow", "magenta", "lightblue"
)
```

# Volatility stress scenarios
## Helper functions
```{r helper_functions}
# convert matrix to column vector
impl_vol_row_to_matrix <- function(vector) {
  # option periods different across cols
  # swap periods different across rows
  return(matrix(as.numeric(vector), nrow = 5, byrow = FALSE))
}

# visualize 5x5 surface plot
plotly_3d <- function(vectors, title, name_z_axis = "implied volatility") {
  surface_plot <- plot_ly()
  n_planes <- 1
  if (!is.vector(vectors)) {
    n_planes <- ncol(vectors)
  }
  colorscales <- c("Rainbow", "YlGnBu", "Hot", "Jet", "Picnic", "Portland", "RdBu", "Reds", "Viridis", "YlOrRd", "Earth")
  for (i in 1:n_planes) {
    if (n_planes == 1) {
      matrix <- impl_vol_row_to_matrix(vectors)
    } else {
      matrix <- impl_vol_row_to_matrix(vectors[, i])
    }

    surface_plot <- surface_plot |> add_surface(
      x = seq(5, 25, 5), y = seq(5, 25, 5), z = matrix,
      contours = list(
        x = list(show = TRUE),
        y = list(show = TRUE),
        z = list(
          show = TRUE,
          usecolormap = TRUE
          # project=list(z=TRUE)
        )
      ),
      opacity = 0.6,
      colorscale = colorscales[i],
      colorbar = list(title = colnames(vectors)[i])
    )
  }

  surface_plot <- surface_plot |>
    layout(
      title = title,
      scene = list(
        xaxis = list(
          title = "option period",
          tickvals = seq(5, 25, 5)
        ),
        yaxis = list(
          title = "swap tenor",
          tickvals = seq(5, 25, 5)
        ),
        zaxis = list(
          title = name_z_axis
        ),
        camera = list(
          eye = list(x = 1.8, y = 1.8, z = -0.5)
        )
      )
    )
  return(surface_plot)
}
```

```{r projection_on_principal_components}
# determine joint quantiles using quantile regressions
determine_joint_quantile <- function(three_dim_obs, quantile_lvl) {
  colnames(three_dim_obs) <- c("pc1", "pc2", "pc3")
  three_dim_obs <- three_dim_obs[rowSums(abs(three_dim_obs) > 10) == 0, ]

  q1 <- quantile(three_dim_obs$pc1, probs = quantile_lvl)
  summary(q1) |> print()
  quantile_regression_1 <- rq(pc2 ~ ., tau = quantile_lvl, data = three_dim_obs |> select(-pc3))
  summary(quantile_regression_1) |> print()
  q2 <- predict(quantile_regression_1, newdata = tibble(pc1 = q1))
  quantile_regression_2 <- rq(pc3 ~ ., tau = quantile_lvl, data = three_dim_obs)
  summary(quantile_regression_2) |> print()
  q3 <- predict(quantile_regression_2, newdata = tibble(pc1 = q1, pc2 = q2))

  # Define the vertical plane x = 2 (for example)
  x <- seq(min(three_dim_obs$pc1), max(three_dim_obs$pc1, q1), length.out = 10)
  y <- seq(min(three_dim_obs$pc2), max(three_dim_obs$pc2, q2), length.out = 10)
  z <- seq(min(three_dim_obs$pc3), max(three_dim_obs$pc3, q3), length.out = 10)
  xy <- meshgrid(x, y)

  plot <- plot_ly(three_dim_obs, x = ~pc1, y = ~pc2, z = ~pc3, type = "scatter3d", name = "observations") |>
    # quantile of pc1
    add_surface(
      x = matrix(q1, nrow = length(y), ncol = length(z)),
      y = matrix(y, nrow = length(y), ncol = length(z)),
      z = matrix(z, nrow = length(y), ncol = length(z), byrow = TRUE),
      opacity = 0.5, showscale = FALSE
    ) |>
    # quantile regression line pc2 ~ pc1
    add_surface(
      x = matrix(x, nrow = length(y), ncol = length(z)),
      y = matrix(coef(quantile_regression_1)[1] + coef(quantile_regression_1)[2] * x, nrow = length(y), ncol = length(z)),
      z = matrix(z, nrow = length(y), ncol = length(z), byrow = TRUE),
      opacity = 0.5, showscale = FALSE
    ) |>
    # quantile regression line pc3 ~ pc2 + pc1
    add_surface(
      x = x,
      y = y,
      z = coef(quantile_regression_2)[1] + coef(quantile_regression_2)[2] * xy$X + coef(quantile_regression_2)[3] * xy$Y,
      opacity = 0.5, showscale = FALSE
    ) |>
    # joint quantile
    add_trace(x = q1, y = q2, z = q3, name = "joint quantile") |>
    layout(
      title = paste0("Joint ", quantile_lvl * 100, "% quantile"),
      xaxis = list(
        title = "pc1 scores",
        range = list(-2, 7)
      ),
      yaxis = list(
        title = "pc2 scores",
        range = list(-1, 2.5)
      ),
      zaxis = list(
        title = "pc3 scores",
        range = list(-1.4, 2.5)
      )
    )

  return(list(
    joint_quantile = c(q1, q2, q3),
    plot = plot,
    three_dim_obs = three_dim_obs
  ))
}

# visualize and annualize principal component analysis results
principal_component_viz <- function(pca, # prcomp class
                                    data,
                                    order_number_max,
                                    with_plots = FALSE,
                                    quantile_lvl = 0.995,
                                    is_weekly = TRUE) {
  pc_scores_copula <- tibble() # annualized principal component projection scores using Copula method
  pc_scores_ar <- tibble() # annualized principal component projection scores using AR(1) method
  month_names <- c("Jan", "Feb", "Mar", "Apr", "Mai", "Jun", "Jul", "Aug", "Sep", "Okt", "Nov", "Dez")

  plots_annualized_copula <- list()
  plots_annualized_ar <- list()

  for (order_number in 1:order_number_max) {
    order_text <- ""
    if (order_number == 1) {
      order_text <- "first"
    } else if (order_number == 2) {
      order_text <- "second"
    } else if (order_number == 3) {
      order_text <- "third"
    }

    if (with_plots) {
      # plot principal component surface of implied volatilities
      plotly_3d(
        pca$rotation[, paste0("PC", order_number)] |> as.numeric(),
        paste0(
          toupper(substr(order_text, 1, 1)), substr(order_text, 2, 100),
          " principal component surface of changes in implied volatilities"
        )
      ) |> print()
    }

    # save projection scores
    df_pc_projection <- tibble(
      timestamp = data$Timestamp,
      projection = NA
    )

    for (i in 1:nrow(data)) {
      df_pc_projection$projection[i] <- sum(
        data[i, -1] * pca$rotation[, paste0("PC", order_number)]
      )
    }
    df_pc_projection <- df_pc_projection |>
      arrange(timestamp) |>
      mutate(
        # week = strftime(timestamp, format = "%V"),
        month = month(timestamp),
        year = year(timestamp)
      )

    if (with_plots) {
      upper_quantile <- quantile(df_pc_projection$projection, quantile_lvl)
      # histogram of all observed principal component scores
      plot <- ggplot(data = df_pc_projection, aes(x = projection)) +
        geom_histogram(aes(y = ..density..)) +
        geom_density(colour = "blue", linewidth = 0.8) +
        geom_vline(aes(xintercept = upper_quantile), linewidth = 0.8, linetype = "dashed", colour = "darkgreen") +
        annotate(geom = "text", x = upper_quantile, y = 0.3, hjust = -0.1, label = formatC(upper_quantile, 4), color = "darkgreen") +
        annotate(
          geom = "text", x = -Inf, y = Inf, hjust = 1, vjust = 1,
          label = TeX(paste0(
            "$\\hat{\\mu}$ = ", formatC(mean(df_pc_projection$projection), digits = 4),
            ", $\\hat{\\sigma}^2$ = ", formatC(var(df_pc_projection$projection), digits = 4)
          ))
        ) +
        labs(
          title = paste0("Histogram of ", order_text, " principal component scores"),
          subtitle = paste0(""),
          x = "principal component score",
          y = "density"
        )
      print(plot)
    }

    # fit normal distribution for principal component scores (monthly frequency)
    pc_scores_distr <- fitdistr(df_pc_projection$projection, "t")$estimate
    print(pc_scores_distr)

    df_pc_projection_wide <- df_pc_projection
    for (i in 1:ifelse(is_weekly, 51, 11)) {
      df_pc_projection_wide <- df_pc_projection_wide |>
        mutate(
          "lag{{i}}" := lag(projection, i)
        )
    }

    df_pc_projection_wide <- df_pc_projection_wide |>
      select(-timestamp, -month, -year) |>
      drop_na()
    colnames(df_pc_projection_wide) <- paste0("lag_", 0:ifelse(is_weekly, 51, 11))

    if (with_plots) {
      # histogram of all observed principal component scores grouped by month
      plot <- ggplot(data = gather(df_pc_projection_wide), aes(x = value)) +
        geom_histogram(aes(y = ..density..)) +
        geom_density(colour = "blue", linewidth = 0.8) +
        facet_wrap(~key, scales = "free") +
        labs(
          title = paste0("Histogram of ", order_text, " principal component scores"),
          subtitle = paste0("Groups by month"),
          x = "principal component score",
          y = "density"
        )
      print(plot)
    }

    nsim <- 100000
    ##############################################################################
    # 1) Annualization: Monthly simulations using Copula #########################
    ##############################################################################
    if (with_plots) {
      # covariance matrix of projection scores
      pc_scores_cov <- df_pc_projection_wide |>
        cor(use = "pairwise.complete.obs") |>
        ggcorrplot() +
        labs(
          title = paste0(
            "Autocorrelation matrix of observed ", ifelse(is_weekly, "weekly", "monthly"),
            " principal component scores"
          )
        ) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
      print(pc_scores_cov)
    }

    set.seed(1)
    # fit 12 dimensional normal copula and simulate from it
    # normcop = normalCopula(dim = ncol(df_pc_projection_wide), dispstr='toep')
    normcop <- tCopula(dim = ncol(df_pc_projection_wide), dispstr = "toep", df.fixed = FALSE)
    fit <- fitCopula(normcop, pobs(as.matrix(df_pc_projection_wide)), method = "mpl")
    print(fit@estimate)
    simulations <- rCopula(nsim, tCopula(fit@estimate[-length(fit@estimate)], df = fit@estimate[length(fit@estimate)], dim = ncol(df_pc_projection_wide), dispstr = "toep"))

    if (with_plots) {
      # 2d marginals of observed scores
      plot_obs <- ggplot(tibble(x = pt((pca$x[1:597, "PC1"] - 0.0004) / 0.0607, df = 3.173), y = pt((pca$x[2:598, "PC1"] - 0.0004) / 0.0607, df = 3.173)), aes(x, y)) +
        geom_point() +
        labs(
          title = "Observations of lag_0 and lag_1 scores",
          subtitle = "Dependence structure of 598 weekly first principal component scores",
          x = TeX("$F_{(\\mu,\\tau,\\nu)}(x_{i})$ of lag_0 observations $x_i$"),
          y = TeX("$F_{(\\mu,\\tau,\\nu)}(x_{i})$ of lag_1 observation $x_i$")
        )
      # 2d marginals of copula simulations
      plot_sim <- ggplot(tibble(x = simulations[1:597, 1], y = simulations[1:597, 2]), aes(x, y)) +
        geom_point() +
        labs(
          title = "Simulation of dependent lag_0 and lag_1 scores",
          subtitle = "Dependence structure of 598 t Copula marginals",
          x = "marginal uniform distribution of lag_0",
          y = "marginal uniform distribution of lag_1"
        )
      print(ggarrange(ggMarginal(plot_obs, type = "density"), ggMarginal(plot_sim, type = "density"), nrow = 1))
      print(paste("Correlation small: ", cor(
        qt(simulations[1:597, 1], df = pc_scores_distr["df"]) * pc_scores_distr["s"] + pc_scores_distr["m"],
        qt(simulations[1:597, 2], df = pc_scores_distr["df"]) * pc_scores_distr["s"] + pc_scores_distr["m"]
      )))
      print(paste("Correlation large: ", cor(
        qt(simulations[, 1], df = pc_scores_distr["df"]) * pc_scores_distr["s"] + pc_scores_distr["m"],
        qt(simulations[, 2], df = pc_scores_distr["df"]) * pc_scores_distr["s"] + pc_scores_distr["m"]
      )))
    }

    # re-transform uniform copula estimates to empirical marginal distribution
    for (colIndex in 1:ncol(simulations)) {
      # simulations[,colIndex] = quantile(ecdf(df_pc_projection_wide[[colIndex]]), simulations[,colIndex])
      # simulations[,colIndex] = qnorm(simulations[,colIndex], mean = pc_scores_distr['mean'], sd = pc_scores_distr['sd'])
      simulations[, colIndex] <- qt(simulations[, colIndex], df = pc_scores_distr["df"]) * pc_scores_distr["s"] + pc_scores_distr["m"]
    }
    simulations <- as_tibble(simulations)
    colnames(simulations) <- colnames(df_pc_projection_wide)

    # 2d marginals of copula simulations
    plot <- ggplot(tibble(x = simulations$lag_0, y = simulations$lag_1), aes(x, y)) +
      geom_point() +
      labs(
        title = "Simulation of dependent lag_0 and lag_1 scores",
        subtitle = "Excerpt of a 52-dimensional t Copula",
        x = "lag_0",
        y = "lag_1"
      )
    print(ggMarginal(plot, type = "density"))

    if (with_plots) {
      # covariance matrix of simulations
      pc_scores_cov <- simulations |>
        cor(use = "pairwise.complete.obs") |>
        ggcorrplot() +
        labs(
          title = paste0("Autocorrelation matrix of observed ", ifelse(is_weekly, "weekly", "monthly"))
        )
      print(pc_scores_cov)

      # histogram of simulated principal component scores grouped by month
      plot <- ggplot(data = gather(simulations), aes(x = value)) +
        geom_histogram(aes(y = ..density..)) +
        geom_density(colour = "blue", linewidth = 0.8) +
        facet_wrap(~key, scales = "free") +
        labs(
          title = paste0("Histogram of ", order_text, " principal component simulated scores"),
          subtitle = paste0("Groups by month"),
          x = "simulation",
          y = "density"
        )
      print(plot)
    }

    # Annualized monthly simulations
    annualized_simulations_copula <- apply(1 + simulations, 1, prod) - 1
    upper_quantile_copula <- quantile(annualized_simulations_copula, quantile_lvl, na.rm = TRUE)
    # remove outliers for histogram
    annualized_simulations_copula_hist <- annualized_simulations_copula[order(annualized_simulations_copula, decreasing = TRUE)[-(1:40)]]
    annualized_simulations_copula_hist <- annualized_simulations_copula_hist[order(annualized_simulations_copula_hist, decreasing = FALSE)[-(1:40)]]
    if (with_plots) {
      plot <- ggplot(data = tibble(value = annualized_simulations_copula_hist), aes(x = value)) +
        geom_histogram(aes(y = ..density..)) +
        # geom_line(aes(y = dnorm(value, mean = mean(annualized_simulations_copula), sd = sqrt(var(annualized_simulations_copula))))) +
        geom_density(colour = "blue", linewidth = 0.8) +
        geom_vline(aes(xintercept = upper_quantile_copula), linewidth = 0.8, linetype = "dashed", colour = "darkgreen") +
        annotate(geom = "text", x = upper_quantile_copula, y = 0.3, hjust = -0.1, label = formatC(upper_quantile_copula, 4), color = "darkgreen") +
        # annotate(geom = "text", x = -Inf, y = Inf, hjust=0, vjust=1,
        #          label = TeX(paste0("$\\hat{\\mu}$ = ", formatC(mean(annualized_simulations_copula), digits=4),
        #             ", $\\hat{\\sigma}^2$ = ", formatC(var(annualized_simulations_copula), digits=4)))) +
        labs(
          title = paste0("Histogram of ", order_text, " principal component annualized simulated scores"),
          subtitle = paste0("Annualization performed using Copula simulations"),
          x = "simulation",
          y = "density"
        )
      print(plot)
      plots_annualized_copula[[order_number]] <- plot
    }

    ##############################################################################
    # 2) Annualization: Monthly simulations using AR(1) process ##################
    ##############################################################################
    if (with_plots) {
      plot_projection <- ggplot() +
        geom_line(data = df_pc_projection, mapping = aes(timestamp, projection)) +
        labs(
          title = paste0("Scores of ", order_text, " principal component"),
          x = "time",
          y = ""
        )
      print(order_text)
      print(plot_projection)
    }

    acf <- acf(df_pc_projection$projection, lag.max = 50, plot = FALSE)
    df_acf <- with(acf, data.frame(lag, acf))

    if (with_plots) {
      conf.level <- 0.95
      ciline <- qnorm((1 - conf.level) / 2) / sqrt(length(df_pc_projection$projection))
      plot_acf <- ggplot(data = df_acf, mapping = aes(x = lag, y = acf)) +
        geom_line() +
        labs(
          title = paste0("Autocorrelation function of ", order_text, " principal component"),
          x = "Lag (weeks)",
          y = ""
        )
    }

    pacf <- pacf(df_pc_projection$projection, lag.max = 50, plot = FALSE)
    pdf_acf <- with(pacf, data.frame(lag, acf))

    if (with_plots) {
      plot_pacf <- ggplot(data = pdf_acf, mapping = aes(x = lag, y = acf)) +
        geom_hline(aes(yintercept = 0)) +
        geom_segment(mapping = aes(xend = lag, yend = 0)) +
        geom_hline(aes(yintercept = ciline), linetype = 2, color = "darkblue") +
        geom_hline(aes(yintercept = -ciline), linetype = 2, color = "darkblue") +
        labs(
          title = paste0("Sample partial autocorrelation function of ", order_text, " principal component"),
          x = "Lag (weeks)",
          y = "Sample partial autocorrelations"
        )
      print(plot_acf / plot_pacf)
    }
    
    # fit AR(1) model to principal component scores time series
    projection_score_time_series <- ar(df_pc_projection$projection, order.max = 1, aic = FALSE)
    sigma <- var(projection_score_time_series$resid[-1]) |> sqrt()
    print(projection_score_time_series)
    print(paste("AR(1)-process: sigma = ", sigma))
    plot_ar1_residuals_time <- ggplot(
      data = tibble(time = df_pc_projection$timestamp[-1], residuals = projection_score_time_series$resid[-1]),
      mapping = aes(x = time, y = residuals)
    ) +
      geom_line() +
      labs(
        title = "Time series plot of residuals of AR(1) process",
        subtitle = paste0(order_text, " principal component"),
        x = "time",
        y = "residual"
      )
    plot_ar1_residuals_hist <- ggplot(
      data = tibble(time = df_pc_projection$timestamp[-1], residuals = projection_score_time_series$resid[-1]),
      mapping = aes(x = residuals)
    ) +
      geom_histogram() +
      geom_rug() +
      labs(
        title = "Histogram of residuals of fitted AR(1) process",
        subtitle = paste0(order_text, " principal component"),
        x = "residual",
        y = "frequency"
      )
    plot_ar1_residuals_qq <- ggplot(tibble(residuals = projection_score_time_series$resid[-1]), aes(sample = residuals)) +
      stat_qq() +
      stat_qq_line() +
      labs(
        title = "QQ-plot: Residuals of fitted AR(1) process",
        subtitle = "Quantiles of residuals vs normal distribution",
        x = "theoretical quantile",
        y = "sample quantile"
      )
    print(plot_ar1_residuals_time | plot_ar1_residuals_hist | plot_ar1_residuals_qq)

    # simulate AR(1) time series and estimate annual rate of change
    set.seed((order_number - 1) * nsim + 1)
    annualized_simulations_ar <- numeric(nsim)
    for (i in 1:nsim) {
      annualized_simulations_ar[i] <- (1 + arima.sim(n = ifelse(is_weekly, 52, 12), model = list(order = c(1, 0, 0), ar = projection_score_time_series$ar), sd = sigma)) |>
        prod()
    }
    annualized_simulations_ar <- annualized_simulations_ar - 1
    upper_quantile_ar <- quantile(annualized_simulations_ar, quantile_lvl, na.rm = TRUE)
    # remove outliers for histogram
    annualized_simulations_ar_hist <- annualized_simulations_ar[order(annualized_simulations_ar, decreasing = TRUE)[-(1:40)]]
    annualized_simulations_ar_hist <- annualized_simulations_ar_hist[order(annualized_simulations_ar_hist, decreasing = FALSE)[-(1:40)]]
    if (with_plots) {
      plot <- ggplot(data = tibble(value = annualized_simulations_ar_hist), aes(x = value)) +
        geom_histogram(aes(y = ..density..)) +
        # geom_line(aes(y = dnorm(value, mean = mean(annualized_simulations_ar), sd = sqrt(var(annualized_simulations_ar))))) +
        geom_density(colour = "blue", linewidth = 0.8) +
        geom_vline(aes(xintercept = as.numeric(upper_quantile_ar)), linewidth = 0.8, linetype = "dashed", colour = "darkgreen") +
        annotate(geom = "text", x = upper_quantile_ar, y = 0.2, hjust = 0, label = formatC(upper_quantile_ar, 4), color = "darkgreen") +
        # annotate(geom = "text", x = -Inf, y = Inf, hjust=0, vjust=1,
        #          label = TeX(paste0("$\\hat{\\mu}$ = ", formatC(mean(annualized_simulations_ar), digits=4),
        #             ", $\\hat{\\sigma}^2$ = ", formatC(var(annualized_simulations_ar), digits=4)))) +
        labs(
          title = paste0("Histogram of ", order_text, " principal component annualized simulated scores"),
          subtitle = paste0("Annualization performed using AR(1) time series simulations"),
          x = "simulation",
          y = "density"
        )
      print(plot)
      plots_annualized_ar[[order_number]] <- plot
    }
    # cbind projections scores to pc_scores
    if (ncol(pc_scores_ar) > 0) {
      pc_scores_copula <- pc_scores_copula |> cbind(annualized_simulations_copula)
      pc_scores_ar <- pc_scores_ar |> cbind(annualized_simulations_ar)
    } else {
      pc_scores_copula <- tibble(annualized_simulations_copula)
      pc_scores_ar <- tibble(annualized_simulations_ar)
    }
  }
  print(pc_scores_ar)
  # Determine joint quantile at level quantile_lvl
  joint_quantile_copula <- determine_joint_quantile(pc_scores_copula, quantile_lvl)
  joint_quantile_ar <- determine_joint_quantile(pc_scores_ar, quantile_lvl)
  print(joint_quantile_copula$plot)
  print(joint_quantile_ar$plot)


  if (with_plots) {
    print((plots_annualized_copula[[1]] | plots_annualized_ar[[1]]) /
      (plots_annualized_copula[[2]] | plots_annualized_ar[[2]]) /
      (plots_annualized_copula[[3]] | plots_annualized_ar[[3]]))
  }

  return(list(
    joint_quantile_copula = joint_quantile_copula, # $joint_quantile,
    joint_quantile_ar = joint_quantile_ar # $joint_quantile
  ))
}
```

```{r function_time_series_plot}
# Function to scale secondary axis
scale_function <- function(x, scale, shift) {
  return((x - shift) * scale)
}

# Function to scale secondary variable values
inv_scale_function <- function(x, scale, shift) {
  return(x / scale + shift)
}

plot_grouped_implied_volatilities <- function(impl_vol, additional_time_series) {
  data <- impl_vol

  time_series_colnames <- c()
  if (!is.null(additional_time_series)) {
    data <- impl_vol |>
      left_join(additional_time_series, by = join_by(Timestamp))

    time_series_colnames <- colnames(additional_time_series[, -1])

    # max of first y axis
    max_y_first <- max(impl_vol[, -1])
    max_y_second <- max(additional_time_series[, -1], na.rm = TRUE) # max of second y axis
    min_y_first <- min(impl_vol[, -1]) # min of first y axis
    min_y_second <- min(additional_time_series[, -1], na.rm = TRUE) # min of second y axis

    # scale and shift variables calculated based on desired mins and maxes
    scale <- (max_y_second - min_y_second) / (max_y_first - min_y_first)
    shift <- mean(min_y_first - min_y_second, max_y_first - max_y_second)
  } else {
    scale <- 1
    shift <- 0
  }

  data_long <- data |>
    pivot_longer(
      cols = -Timestamp,
      names_to = "swaption",
      values_to = "value"
    )

  plt <- ggplot() +
    geom_line(
      data = data_long |> filter(!(swaption %in% time_series_colnames)),
      mapping = aes(x = Timestamp, y = value, color = swaption), linewidth = 0.6
    ) +
    # annotate(geom = "rect", xmin = ymd("2024-07-01"), xmax = ymd("2024-07-15"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2) +
    # annotate(geom = "rect", xmin = ymd("2019-12-01"), xmax = ymd("2019-12-31"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2) +
    # annotate(geom = "rect", xmin = ymd("2019-05-01"), xmax = ymd("2019-05-30"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2) +
    annotate(geom = "rect", xmin = ymd("2015-05-01"), xmax = ymd("2015-06-01"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2) +
    annotate(geom = "rect", xmin = ymd("2020-03-01"), xmax = ymd("2020-04-01"), ymin = -Inf, ymax = Inf, fill = colors["green"], alpha = 0.2)

  if (!is.null(additional_time_series)) {
    plt <- plt +
      geom_line(
        data = data_long |> filter(swaption %in% time_series_colnames),
        mapping = aes(x = Timestamp, y = inv_scale_function(value, scale, shift), colour = "interest_rate"),
        color = "gray20"
      ) +
      scale_y_continuous(
        name = "implied volatility",
        # limits = c(min_y_first, max_y_first),
        sec.axis = sec_axis(
          ~ scale_function(., scale, shift),
          name = "interest rate",
          labels = waiver()
        )
      )
  }
  plt <- plt +
    scale_color_manual(values = c("20YX5Y" = "red", "interest_rate" = "gray20")) +
    labs(
      title = "Implied volatility for swaptions with same option period",
      subtitle = paste0("Swaptions from 2/7/2013 to 9/9/2024"),
      x = "date"
    ) +
    theme(
      axis.title.y = element_markdown(color = "black"),
      axis.title.y.right = element_markdown(color = colors["gray20"])
    )
  return(plt)
}
```

## Observations: Daily implied volatilities
```{r load_data}
set.seed(1)

swaption_volatilities_d <- read_csv("../data_raw/swaption_volatilities.csv")
swaption_volatilities_d$Timestamp <- mdy(swaption_volatilities_d$Timestamp)

# Count NA rows
paste0(
  "Number of NA rows: ",
  swaption_volatilities_d |> count() |> as.numeric() -
    swaption_volatilities_d |> drop_na() |> count() |> as.numeric()
)
swaption_volatilities_d <- swaption_volatilities_d |>
  drop_na() |>
  filter(Timestamp != ymd("2015-05-04"))
# 24 rows in April 2015 contain missing implied volatilities. These observations are removed.

swaption_volatilities_d |>
  select(-Timestamp) |>
  cor(use = "pairwise.complete.obs") |>
  ggcorrplot() +
  labs(
    title = paste0("Correlation of implied swaption volatilities"),
    subtitle = "Daily data from 2/7/2013 to 9/9/2024"
  )

swaption_volatilities_d_corr <- swaption_volatilities_d |>
  select(-Timestamp) |>
  cor(use = "pairwise.complete.obs")

# ggplot of pairwise correlations in swaption volatilities
swaption_volatilities_d_corr <-
  tibble(
    value = swaption_volatilities_d_corr[upper.tri(swaption_volatilities_d_corr)]
  )

ggplot(data = gather(swaption_volatilities_d_corr), aes(x = value, y = factor(""))) +
  geom_violin() +
  geom_boxplot(width = 0.4) +
  labs(
    title = paste0("Pairwise correlations of implied swaption volatilities"),
    subtitle = paste0("daily historical data of 25 swaptions from 2/7/2013 to 9/9/2024"),
    x = "Pearson correlation",
    y = ""
  )

# relative change in daily swaption volatilities
n <- nrow(swaption_volatilities_d)
swaption_volatilities_d_rel <- (swaption_volatilities_d[1:(n - 1), -1] /
  swaption_volatilities_d[2:n, -1] - 1) |>
  cbind(Timestamp = swaption_volatilities_d$Timestamp[1:(n - 1)])

# relative change in monthly swaption volatilities
swaption_volatilities_m <- swaption_volatilities_d |>
  group_by(month = lubridate::floor_date(Timestamp, "month")) |>
  filter_period(.period = "1 month", Timestamp == first(Timestamp)) |>
  ungroup() |>
  select(-month)
n <- nrow(swaption_volatilities_m)
swaption_volatilities_m_rel <- (swaption_volatilities_m[1:(n - 1), -1] /
  swaption_volatilities_m[2:n, -1] - 1) |>
  cbind(Timestamp = swaption_volatilities_m$Timestamp[1:(n - 1)])

swaption_volatilities_by_date_and_interval <- function(impl_vol_daily, start_date, interval) {
  start_date_year <- year(start_date)
  start_date_month <- month(start_date)
  start_date_day <- day(start_date)

  swaption_volatilities_interval <- NULL
  if (interval == "day") {
    swaption_volatilities_interval <- impl_vol_daily
  } else if (interval == "week") {
    swaption_volatilities_interval <- impl_vol_daily |>
      mutate(Timestamp_shifted = Timestamp - days(7) + 1) |>
      group_by(week = lubridate::floor_date(Timestamp_shifted, "week")) |>
      filter_period(.period = "1 week", .date_var = Timestamp_shifted, Timestamp_shifted == last(Timestamp_shifted)) |>
      ungroup() |>
      select(-c(Timestamp_shifted, week))
  } else if (interval == "month") {
    swaption_volatilities_interval <- impl_vol_daily |>
      mutate(Timestamp_shifted = Timestamp - days(start_date_day) + 1) |>
      group_by(month = lubridate::floor_date(Timestamp_shifted, "month")) |>
      filter_period(.period = "1 month", .date_var = Timestamp_shifted, Timestamp_shifted == last(Timestamp_shifted)) |>
      ungroup() |>
      select(-c(Timestamp_shifted, month))
  } else if (interval == "year") {
    swaption_volatilities_interval <- impl_vol_daily |>
      mutate(Timestamp_shifted = Timestamp - months(start_date_month) - days(start_date_day) + 1) |>
      group_by(year = lubridate::floor_date(Timestamp_shifted, "year")) |>
      filter_period(.period = "1 year", .date_var = Timestamp_shifted, Timestamp_shifted == last(Timestamp_shifted)) |>
      ungroup() |>
      select(-c(Timestamp_shifted, year))
  }
  print(swaption_volatilities_interval)
  # relative change in swaption implied volatilities
  n <- nrow(swaption_volatilities_interval)
  swaption_volatilities_interval_rel <- (swaption_volatilities_interval[1:(n - 1), -1] /
    swaption_volatilities_interval[2:n, -1] - 1) |>
    cbind(Timestamp = swaption_volatilities_interval$Timestamp[1:(n - 1)]) |>
    arrange(Timestamp) |>
    select(Timestamp, everything()) # move Timestamp column to first position

  return(swaption_volatilities_interval_rel)
}

# summary table
swaption_volatilities_d.summary <- swaption_volatilities_d %>%
  pivot_longer(
    cols = -Timestamp,
    names_to = "name",
    values_to = "value"
  ) %>%
  group_by(name) %>%
  summarize(
    min = min(value, na.rm = TRUE),
    quantile1 = quantile(value, 0.01, na.rm = TRUE),
    quantile25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    mean = mean(value, na.rm = TRUE),
    quantile75 = quantile(value, 0.75, na.rm = TRUE),
    quantile99 = quantile(value, 0.99, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  )
pander(swaption_volatilities_d.summary)

# replace outlier by mean of adjacent values
swaption_volatilities_d[swaption_volatilities_d$Timestamp == ymd("2022-11-07"), "25YX5Y"] <-
  0.5 * (swaption_volatilities_d[swaption_volatilities_d$Timestamp == ymd("2022-11-04"), "25YX5Y"] |> as.numeric() +
    swaption_volatilities_d[swaption_volatilities_d$Timestamp == ymd("2022-11-08"), "25YX5Y"] |> as.numeric())

plotly_3d(
  apply(swaption_volatilities_d[, -1], 2, mean),
  "Average implied volatility surface"
)
plotly_3d(
  apply(swaption_volatilities_d[, -1], 2, var) |> sqrt(),
  "Daily standard deviation of implied volatilities"
)

interest_yield <- read_csv("../data_raw/yield_curves_euro_area_aaa.csv")
interest_yield$Timestamp <- ymd(interest_yield$Timestamp)

filters <- c("^5YX", "^10YX", "^15YX", "^20YX", "^25YX")
for (filter in filters) {
  col_filter_indicator <- attr(regexpr(filter, colnames(swaption_volatilities_d)), "match.length") > 0
  col_filter_indicator[1] <- TRUE
  data <- swaption_volatilities_d[, col_filter_indicator]
  additional_time_series <- NULL
  if (filter == "^5YX") {
    additional_time_series <- interest_yield[, c("Timestamp", "yield_5y")]
  } else if (filter == "^10YX") {
    additional_time_series <- interest_yield[, c("Timestamp", "yield_10y")]
  } else if (filter == "^20YX") {
    additional_time_series <- interest_yield[, c("Timestamp", "yield_20y")]
  }
  plot <- plot_grouped_implied_volatilities(data, additional_time_series)
  print(plot)
}

swaption_volatilities_d_scaled <- scale(swaption_volatilities_d[, -1], center = TRUE, scale = FALSE)
swaption_volatilities_d_scaled <- tibble(Timestamp = swaption_volatilities_d$Timestamp) |> cbind(swaption_volatilities_d_scaled)
```

```{r swaptions_implied_volatility_surfaces}
# plot 5x5 implied volatility surface
# data: tibble with implied volatility data, timestamp column at position one
# title: title of plot
plot_surface <- function(data, title) {
  points_in_time <- 1:nrow(data)

  aval <- list()
  for (point_in_time in points_in_time) {
    aval[[point_in_time]] <- list(
      visible = FALSE,
      name = paste0("Date: ", data[point_in_time, 1])
    )
  }
  aval[3][[1]]$visible <- TRUE

  steps <- list()
  fig <- plot_ly()
  for (i in points_in_time) {
    swaption_volatilities_matrix <- impl_vol_row_to_matrix(as.numeric(data[i, -1]))
    fig <- add_surface(fig,
      x = seq(5, 25, 5), y = seq(5, 25, 5), z = swaption_volatilities_matrix,
      name = aval[i][[1]]$name,
      visible = aval[i][[1]]$visible,
      showscale = FALSE,
      contours = list(
        x = list(show = TRUE),
        y = list(show = TRUE),
        z = list(
          show = TRUE,
          usecolormap = TRUE
        )
      )
    )
    step <- list(
      args = list("visible", rep(FALSE, length(aval))),
      method = "restyle",
      label = data[[i, 1]]
    )
    step$args[[2]][i] <- TRUE
    steps[[i]] <- step
  }

  fig <- fig |>
    layout(
      title = title,
      scene = list(
        xaxis = list(
          title = "option period",
          tickvals = seq(5, 25, 5)
        ),
        yaxis = list(
          title = "swap period",
          tickvals = seq(5, 25, 5)
        ),
        zaxis = list(
          title = "implied volatility",
          range = quantile(data[, -1], probs = c(0.01, 0.99), na.rm = TRUE)
        ),
        camera = list(
          eye = list(x = 1.8, y = 1.8, z = -0.5)
        )
      ),
      sliders = list(list(
        active = 2,
        currentvalue = list(prefix = "Date: "),
        steps = steps
      ))
    )

  return(fig)
}

plot_surface(swaption_volatilities_d, "Daily implied normal volatilities")
```
The range and shape of implied volatiltity surfaces differ in the course of time. Four exemplary surfaces on four different dates present variations of the volatiltity surfaces over time:

* **2024-07-11**: The implied volatilities range from $48.6$ to $83.9$. The surface is convex. The value is the lowest for a 25YX25 swaption and highest for a 5YX5Y swaption. The shorter the swap or option period, the higher the implied volatility. The magnitude of change in implied volatilities is greater for a change in the option period than for a change in the swap period.

* **2019-12-03**: The implied volatilities range from $43.6$ to $56.1$. The surface is concave. The 25YX25Y swaption has the lowest value whereas the 10YX10Y has the highest value. For option periods greater than five years, the implied volatilities decrease in the swap and option periods. The implied volatilities for swaptions with option period five are smaller compared to swaptions with option period ten and same swap period. In particular, the volatilitiy surface has a remarkable drop in implied volatilities for the 5YX5Y swaption.

* **2019-05-28**: The implied volatilities range from $42.1$ to $53.9$. The surface is concave. The 25YX25Y swaption has the lowest implied volatility. The 10YX5Y swaption exhibits the highest implied volatility. For option periods greater than five years, the implied volatilities decrease in the swap and option periods. The implied volatilities for swaptions with option period five are smaller compared to swaptions with option period ten and same swap period.

* **2015-01-20**: The implied volatilities range from $41.4$ to $72.0$. The values increase for decreasing swap and option periods with variable magnitude. The implied volatility of $64.3$ for 5YX5Y swaptions is clearly smaller than for 5YX10Y, 10YX5Y or 10YX10Y options.

### Observations: Monthly rate of changes
```{r}
swaption_volatilities_m_rel <- swaption_volatilities_by_date_and_interval(swaption_volatilities_d, start_date = ymd("2013-02-01"), "month")

plot_surface(swaption_volatilities_m_rel, "Monthly rate of change in implied normal volatilities")

plotly_3d(
  apply(swaption_volatilities_m_rel[, -1], 2, mean),
  "Average surface of monthly changes in implied volatility"
)
plotly_3d(
  apply(swaption_volatilities_m_rel[, -1], 2, var) |> sqrt(),
  "Standard deviation surface of monthly changes in implied volatilities"
)
```

### Observations: Weekly rate of changes
```{r}
swaption_volatilities_w_rel <- swaption_volatilities_by_date_and_interval(swaption_volatilities_d, start_date = ymd("2013-02-01"), "week")

plotly_3d(
  apply(swaption_volatilities_w_rel[, -1], 2, mean),
  "Average surface of weekly changes in implied volatility"
)
plotly_3d(
  apply(swaption_volatilities_w_rel[, -1], 2, var) |> sqrt(),
  "Standard deviation surface of weekly changes in implied volatilities"
)
```

**Observed annual change in implied volatilities**
```{r, observed_annual_change}
start_dates <- (swaption_volatilities_d |> arrange(Timestamp))$Timestamp[1:237]

swaption_volatilities_y_rel_stack <- tibble()
for (i in 1:length(start_dates)) {
  swaption_volatilities_y_rel_stack <- swaption_volatilities_y_rel_stack |> rbind(swaption_volatilities_by_date_and_interval(swaption_volatilities_d, start_dates[i], "year"))
}
tmp <- swaption_volatilities_y_rel_stack |>
  rowwise() |>
  mutate(mean = mean(c_across(where(is.numeric)))) |>
  ungroup() |>
  arrange(mean) |>
  select(-mean)

impl_vol_change_untransformed <- tmp[ceiling(0.995 * nrow(tmp)), -1]
plotly_3d(
  impl_vol_change_untransformed |> as.numeric(),
  "99.5% quantile of observed annual change in implied volatilities"
)
```
The plot above shows the empirical 99.5% quantile of the average annual change in implied volatilities among all available reference dates.
